\large{\bf Large effects and the infinitesimal model }

### 2023-12-02 review-responses.tex

Thanks for the careful reading and useful comments.

\begin{point}{}
    Paragraph starting ``This suggests exploring whether stable distributions..'' \revref:
The questions as stated: e.g., whether stable distributions can ``stand in'' for a Gaussian distribution in the infinitesimal model or whether the results of Barton et al ``carry over'' when the sum of effect sizes follows a stable law are quite vague and allow for multiple interpretations. Ideally, in order to address whether Barton et al's results (do not) carry over, one would like to at least show that with a power law distribution of effect sizes, the ``Mendelian noise term'', i.e., the difference between the offspring trait value and the midparent value, is (not) independent of the midparent value in the limit of a very large number of loci. And indeed this is what the authors address
somewhat non-rigorously (and for the case of a Cauchy distribution of effect sizes) in section 5 and using simulations : see e.g., figures 4C and 4f, figures 6C and 6F. Unsurprisingly, these simulations show that with e.g., Cauchy distribution of effect sizes, the non-independence between the segregation/noise term and the midparent trait value does not hold. However, a lot of the mathematical analysis in the paper (sections 3 and 4) explores when the distribution of trait values in a population is (not) Gaussian, which is a completely different question.
\end{point}

\reply{
    We don't actually agree with this characterization --
    sections 3 and 4 explore the consequences of different distributions
    of the Mendelian sampling term (which derives from the sum of effects),
    not the population distribution.
    However, the assumptions and motivation were not nearly as clear as they could have been
    in the previous draft, and this is hopefully more clear now --
    see responses to points below.
}

@@

\begin{point}{}
    I think it is also not correct to claim that ``Since models can be easily set up for which trait distributions are not Gaussian, the implication of this is that for such models, independence of the Mendelian sampling term is not likely a good assumption.'' \revref. As far as I understand, a non-Gaussian distribution of trait values can arise quite easily under the infinitesimal model (where the Mendelian segregation term is independent of the midparent value) e.g, due to strong or non-Gaussian selection or migration or some combination of the two (see, e.g., Figure 1 in Barton et al).
\end{point}

\reply{
    Thanks -- we've changed this to ``since models can be easily set up for which the distribution
    of Mendelian sampling terms is not Gaussian, \ldots''
    and provided additional context \revref.
}

\begin{point}{}
    Section 4, first paragraph \revref: I don't think the third assumption (that trait values in the population follow a Gaussian distribution) is a component of the ``infinitesimal model''. I think Turelli (2017) is also quite clear about this stating that a Gaussian distribution of trait values should only emerge for Gaussian or very weak selection on the population.
\end{point}

\reply{
    We agree that \citet{turelli2017commentary} is clear about when the Gaussian emerges.
    It also seems quite clear to us that what we say in this paragraph agrees
    with \citet{turelli2017commentary}, who says in his Abstract that
    \begin{quote}
    This commentary distinguishes three nested approximations, referred to as “infinitesimal genetics,” “Gaussian descendants” and “Gaussian population,” each plausibly called “the infinitesimal model.” 
    \end{quote} 
    We think that the underlying issue here may be about motivation for the section.
    To make the motivation, and the point about the Gaussian population assumption more clear,
    we've added some explanatory text \revref.
    Also note that this section does \emph{not} make a Gaussian-population (or, non-Gaussian population) assumption;
    to make this more clear we've changed the phrase later in this paragraph
    from ``non-Gaussian model'' to ``model with non-Gaussian Mendelian sampling terms''.
}

@@

\begin{point}{}
More generally, I found it a bit hard to follow the overall logical flow of section 4.1, especially once the authors launch into an exploration of the ``reproduction'' and ``noise'' terms (pages 9 and 10).
    It would be useful to clarify what the biological/intuitive meaning of the noise and reproduction terms at the outset. It would also be useful to state (at the beginning) what the main goal of these explorations is: (is the goal to identify when both terms ``have well-posed limits independent of the other'' as stated at \llname{limit_remark}), and also to summarise (at the end) to what extent one can do this in all generality.
\end{point}

\reply{
    Good suggestion; we've added more clarifying points;
    see \llname{clarification1}
}

### 2023-12-02 writeup.tex
peter edits

As observed in \cite{turelli2017commentary}, \revpoint{1}{8}
there are at least three interpretations of what is called ``the infinitesimal model''.  The maximalist of these, what \cite{turelli2017commentary} calls the Gaussian-population approximation, assumes a large number of small-effect loci,  parental trait-independent Gaussian Mendelian sampling terms  (as rigorously justified in \cite{barton2017infinitesimal,barton2022infinitesimal}), and, moreover, that  population trait values are in deterministic mutation-selection balance, with a Gaussian empirical distribution
(\ie the \emph{fraction} of the population with trait value in $[z,z+dz)$
is approximately $\frac{1}{2\pi\sigma^{2}} e^{-(z-m)^{2}/(2\sigma^{2})} dz$
for suitable choices of the empirical mean and variance $m$ and $\sigma^{2}$).
Leaving aside
% -- as the Gaussian-population approximation frequently does --
the question of whether or not some combination of effect sizes, dominance, and epistasis can justify non-Gaussian, trait-independent,  Mendelian sampling terms, we might reasonably ask when (or if)
deterministic mutation-selection balance is possible in a non-Gaussian model
that retains independence of the Mendelian sampling terms.
(Indeed, above we have put substantial restrictions on what models of inheritance
might lead to such models.)
We might also ask whether such a model
might still reasonably approximate an additive model with non-Gaussian effects,
and for the purpose of investigating this with simulations,
some calculations will be helpful.

Concretely, consider a  Moran process where an individual whose trait value is $Z$ dies at rate $\mu(Z)$
and is replaced by the offspring of two randomly chosen individuals,
i.e., a new individual with trait
\begin{align*}
  \frac{Z^{1} +Z^{2}}{2} + R ,
\end{align*}
where $Z^{1}$ and $Z^{2}$ are independent draws from the empirical population distribution
and $R$ is an independent draw from the Mendelian sampling distribution
(independent of $Z^{1}$ and $Z^{2}$, with distribution to be specified). 
Mutation-selection balance demands that that
the distribution of individuals removed by death
and the distribution of individuals added by birth are equal,
\ie that $(Z^{1} +Z^{2})/2 + R$ is equal in distribution
to a $\mu$-weighted choice of $Z$.
% In other words, for all bounded and continuous functions $f$,
% \begin{equation} \label{eqn:MSB_condition}
%     \E\left[ f\left(\frac{Z^{1} +Z^{2}}{2} + R \right) \right] = \frac{\E\left[ f(Z) \mu(Z) \right]}{\E[\mu(Z)]},
% \end{equation}
% where the expectation is with respect to the joint law of the population empirical distribution (the law of $Z$, $Z^{1}$ and $Z^{2}$), and $R$.
% 
% Since the set of functions $\mathcal{S} = \{e^{i \theta x} : \theta \in \R\}$ is separating (see \eg \S 3.4 in \cite{Ethier+Kurtz1986}), it suffices to show that \eqref{eqn:MSB_condition} holds for $f \in \mathcal{S}$, \ie
Taking the characteristic function of both these distributions,
deterministic mutation-selection balance is therefore equivalent to
\begin{align} \label{eqn:MSB_CF}
    \frac{\E\left[ e^{i \theta Z}\mu(Z) \right]}{\E[\mu(Z)]}
    =
    \E\left[ e^{i\theta\left(\frac{Z^{1} +Z^{2}}{2} + R \right)}\right] .
\end{align}
Now, we assumed that $Z^{1}$, $Z^{2}$, and $R$ are independent, so we can write \eqref{eqn:MSB_CF} as 
\begin{align*}
    \frac{\E\left[ e^{i\theta Z} \mu(Z) \right]}{\E[\mu(Z)]} &= \E\left[ e^{i\theta\left(\frac{Z^{1} +Z^{2}}{2} + R \right)}\right]\\
	% &= \E\left[ e^{i\frac{\theta}{2} Z^{1}}e^{i\frac{\theta}{2} Z^{2}}e^{\i\theta R}\right]\\
	&= \E\left[e^{i\frac{\theta}{2} Z^{1}}\right]\E\left[e^{i\frac{\theta}{2} Z^{2}}\right]\E\left[e^{i\theta R}\right]\\
	&= \phi\left(\frac{\theta}{2}\right)^{2} \Phi(\theta),
\end{align*}
where
$ \phi(\theta) = \mathbb{E}\left[e^{i\theta Z^{1}}\right] = \mathbb{E}\left[e^{i\theta Z^{2}}\right] $
and
$ \Phi(\theta) = \mathbb{E}\left[e^{i\theta R}\right] $
are the characteristic functions of $Z^{j}$ and $R$, respectively.  In what follows, we will explore some non-Gaussian solutions $\phi(\theta)$ to \eqref{eqn:MSB_CF}.

@@

First, consider the case without selection, so $\mu$ is a constant
which we may take equal to 1.  Then, our condition on the characteristic function simplifies to
\begin{equation}\label{eq:CF_neutral}
	{\textstyle \phi\left(\frac{\theta}{2}\right)^{2}} \Phi(\theta) %e^{-\frac{\eta^{2}\theta^{2}}{2}} 
	= \phi(\theta) ,
\end{equation}
\ie $\phi(\theta)$ is a fixed point of $F$, where
 \[
	F(\phi)(\theta) = {\textstyle \phi\left(\frac{\theta}{2}\right)^{2}} \Phi(\theta),
\]
whereas any fixed point must be the limit of an arbitrary characteristic function, $\varphi(\theta)$, under repeated applications of $F$,
\ie \revpoint{2}{7}
\begin{align*}
	\phi(\theta) 
    &= \lim_{n \to \infty} F^{(n)}(\varphi)(\theta) \\
    &= \lim_{n \to \infty} \varphi\left(\frac{\theta}{2^{n}}\right)^{2^{n}}\prod_{k=0}^{n-1}\Phi\left(\frac{\theta}{2^{k}}\right)^{2^{k}} .
\end{align*}
% Iterating $F$ $n$ times, we get
% \[
% 	F^{(n)}(\varphi)(\theta) =  {\textstyle \varphi\left(\frac{\theta}{2^{n}}\right)^{2^{n}}}\prod_{k=0}^{n-1} {\textstyle \Phi\left(\frac{\theta}{2^{k}}\right)^{2^{k}}},
% \]
% so that
% \[
% 	\phi(\theta) = \lim_{n \to \infty} \varphi\left(\frac{\theta}{2^{n}}\right)^{2^{n}}\prod_{k=0}^{n-1}\Phi\left(\frac{\theta}{2^{k}}\right)^{2^{k}}
% \]
is a fixed point whenever this limit exists.

@@

The crucial step in the derivation of the infinitesimal model in \cite{barton2017infinitesimal} is to show that, following the arguments in \cite{Fisher1918}, knowing an individual's trait value gives vanishingly little information about the contribution of any given locus as $M \to \infty$, which in turn allows one to show that the within-family Mendelian segregation variance is independent of the parental traits. Here, we will consider how conditioning an individual's trait value affects the law of a single allelic contribution in the Cauchy setting described above. A similar argument can be made for any $\alpha$-stable distribution, but the lack of explicit probability density functions for generic $\alpha \in (0,2)$ adds complexity to the argument without adding any insight.

@@

\begin{figure}
    \begin{center}
        \includegraphics{sims/neutral_seg_noise}
    \end{center}
    \caption{
        The distribution of Mendelian sampling terms
        with Gaussian \textbf{(A-C)} and Cauchy \textbf{(D-F)} effect size distributions,
        from the same simulations shown in Figure~\ref{fig:trait_distrns},
        for all individuals born during the last 1,000 time units of the simulation.
        \textbf{(A,D)} Mendelian noise (offspring trait minus midparent) plotted against midparent.
        \textbf{(B,E)} Histograms of the Mendelian sampling term.
        \textbf{(B,E)} Enrichment in quantiles of the Mendelian sampling term conditional on midparent value.
        To compute the latter,
        let $q_0, q_2, q_4, \ldots, q_{98}, q_{100}$ divide the distribution of the Mendelian sampling term
        into 50 bins with roughly equal numbers in each (i.e., the quantiles),
        let $p_k(a,b)$ denote the proportion of the offspring of parents with midparent value
        is in $[a, b)$ whose Mendelian sampling term is in $[q_{k-1}, q_k)$,
        and let $p_k$ be the proportion of all offspring with Mendelian noise in $[q_{k-1}, q_k)$
        (so, $p_k \approx 0.02$).
        Then, each line shows $p_k(a,b)/p_k$
        plotted against the midpoint of the corresponding quantile bin $(q_{k-1},q_k)$,
        for $(a,b)$ chosen to span the three segments of 5\% of
        ``low'', ``medium'', and ``high'' midparents, i.e.:
        midparents between (red) the 7.5\% and 12.5\% quantiles,
        (black) the 47.5\% and 52.5\% quantiles, and
        (green) the 87.5\% and 92.5\% quantiles of midparent value.
        \label{fig:seg_noise}
    }
\end{figure}

@@

The infinitesimal model is a powerful and tractable model of trait inheritance,
but is polygenic to a fault,
excluding the possibility of loci of large effect
(that are often seen in practice).
In this paper,
we have looked at the possibility of using a broader range of central limit theorems
to extend the infinitesimal model to traits for which 
effect size distributions fall in the domain of attraction of the $\alpha$-stable distributions
(i.e., a larger proportion of variance
is explained by a smaller number of loci).
Indeed, we found evidence that this is true of many real traits,
although our relatively uncritical use of GWAS estimates makes this evidence suggestive at best.
Next, we show that in fact the infinitesimal model \emph{cannot} describe non-Gaussian distributions
\comment{XXX TODO}.
A recurring theme in the rest of the paper is
that just as the infinitesimal model is useful precisely because it ignores the genetic details,
it cannot apply once some genetic details are important
(i.e., there are loci of large effect).

>>>>

As observed in \cite{turelli2017commentary},
there are at least three interpretations of what is called ``the infinitesimal model''. 
The maximalist of these, what \cite{turelli2017commentary} calls the Gaussian-population approximation,
assumes parental trait-independent Gaussian Mendelian sampling terms,
and, moreover, that  population trait values are in deterministic mutation-selection balance, with a Gaussian empirical distribution
(\ie the \emph{fraction} of the population with trait value in $[z,z+dz)$
is approximately $\frac{1}{2\pi\sigma^{2}} e^{-(z-m)^{2}/(2\sigma^{2})} dz$
for suitable choices of the empirical mean and variance $m$ and $\sigma^{2}$).
As also discussed in \citet{turelli2017commentary}, \revpoint{1}{8}
this assumption is rather fragile --
a Gaussian distribution for the population only emerges under weak or Gaussian selection.
However, the model has been used in influential work.
In principle, another factor that could produce a non-Gaussian population distribution
is if the Mendelian sampling terms are not Gaussian, as we consider here --
so, it is interesting to look for other situations where the population distribution is analytically tractable.

Leaving aside 
% -- as the Gaussian-population approximation frequently does --
the question of whether or not some combination of effect sizes, dominance, and epistasis can justify non-Gaussian, trait-independent,  Mendelian sampling terms, we might reasonably ask when (or if)
deterministic mutation-selection balance is possible in a model with non-Gaussian Mendelian sampling terms
that retains independence of these terms.
(Indeed, above we have put substantial restrictions on what models of inheritance
might lead to such models.)
We might also ask whether such a model
might still reasonably approximate an additive model with non-Gaussian effects,
and for the purpose of investigating this with simulations,
some calculations will be helpful.

Concretely, consider a  Moran process where an individual whose trait value is $Z$ dies at rate $\mu(Z)$
and is replaced by the offspring of two randomly chosen individuals,
i.e., a new individual with trait
\begin{align*}
  \frac{Z^{1} +Z^{2}}{2} + R ,
\end{align*}
where $Z^{1}$ and $Z^{2}$ are independent draws from the empirical population distribution
and $R$ is an independent draw from the Mendelian sampling distribution
(independent of $Z^{1}$ and $Z^{2}$, with distribution to be specified). 
Mutation-selection balance demands that that
the distribution of individuals removed by death
and the distribution of individuals added by birth are equal,
\ie that $(Z^{1} +Z^{2})/2 + R$ is equal in distribution
to a $\mu$-weighted choice of $Z$.
% In other words, for all bounded and continuous functions $f$,
% \begin{equation} \label{eqn:MSB_condition}
%     \E\left[ f\left(\frac{Z^{1} +Z^{2}}{2} + R \right) \right] = \frac{\E\left[ f(Z) \mu(Z) \right]}{\E[\mu(Z)]},
% \end{equation}
% where the expectation is with respect to the joint law of the population empirical distribution (the law of $Z$, $Z^{1}$ and $Z^{2}$), and $R$.
% 
% Since the set of functions $\mathcal{S} = \{e^{i \theta x} : \theta \in \R\}$ is separating (see \eg \S 3.4 in \cite{Ethier+Kurtz1986}), it suffices to show that \eqref{eqn:MSB_condition} holds for $f \in \mathcal{S}$, \ie
Taking the characteristic function of both these distributions,
deterministic mutation-selection balance is therefore equivalent to
\begin{align} \label{eqn:MSB_CF}
    \frac{\E\left[ e^{i \theta Z}\mu(Z) \right]}{\E[\mu(Z)]}
    =
    \E\left[ e^{i\theta\left(\frac{Z^{1} +Z^{2}}{2} + R \right)}\right] .
\end{align}
Now, we assumed that $Z^{1}$, $Z^{2}$, and $R$ are independent, so we can write \eqref{eqn:MSB_CF} as 
\begin{align*}
    \frac{\E\left[ e^{i\theta Z} \mu(Z) \right]}{\E[\mu(Z)]} &= \E\left[ e^{i\theta\left(\frac{Z^{1} +Z^{2}}{2} + R \right)}\right]\\
	% &= \E\left[ e^{i\frac{\theta}{2} Z^{1}}e^{i\frac{\theta}{2} Z^{2}}e^{\i\theta R}\right]\\
	&= \E\left[e^{i\frac{\theta}{2} Z^{1}}\right]\E\left[e^{i\frac{\theta}{2} Z^{2}}\right]\E\left[e^{i\theta R}\right]\\
	&= \phi\left(\frac{\theta}{2}\right)^{2} \Phi(\theta),
\end{align*}
where
$ \phi(\theta) = \mathbb{E}\left[e^{i\theta Z^{1}}\right] = \mathbb{E}\left[e^{i\theta Z^{2}}\right] $
and
$ \Phi(\theta) = \mathbb{E}\left[e^{i\theta R}\right] $
are the characteristic functions of $Z^{j}$ and $R$, respectively. 
In what follows, we will explore some non-Gaussian solutions $\phi(\theta)$ to equation \eqref{eqn:MSB_CF}:

@@

First, consider the case without selection, so $\mu$ is a constant
which we may take equal to 1. 
(To remind the reader, our goal is therefore to describe \llabel{clarification1}
what forms the population distribution might take in a neutral model
if we assume only that the Mendelian sampling terms are independent of parental values,
and nothing else about the distribution of these terms or the initial population distribution.) 
Then, our condition on the characteristic function simplifies to
\begin{equation}\label{eq:CF_neutral}
	{\textstyle \phi\left(\frac{\theta}{2}\right)^{2}} \Phi(\theta) %e^{-\frac{\eta^{2}\theta^{2}}{2}} 
	= \phi(\theta) ,
\end{equation}
\ie $\phi(\theta)$ is a fixed point of $F$, where
 \[
	F(\phi)(\theta) = {\textstyle \phi\left(\frac{\theta}{2}\right)^{2}} \Phi(\theta),
\]
whereas any fixed point must be the limit of an arbitrary characteristic function, $\varphi(\theta)$, under repeated applications of $F$,
\ie \revpoint{2}{7}
\begin{align*}
	\phi(\theta) 
    &= \lim_{n \to \infty} F^{(n)}(\varphi)(\theta) \\
    &= \lim_{n \to \infty} \varphi\left(\frac{\theta}{2^{n}}\right)^{2^{n}}\prod_{k=0}^{n-1}\Phi\left(\frac{\theta}{2^{k}}\right)^{2^{k}} .
\end{align*}
% Iterating $F$ $n$ times, we get
% \[
% 	F^{(n)}(\varphi)(\theta) =  {\textstyle \varphi\left(\frac{\theta}{2^{n}}\right)^{2^{n}}}\prod_{k=0}^{n-1} {\textstyle \Phi\left(\frac{\theta}{2^{k}}\right)^{2^{k}}},
% \]
% so that
% \[
% 	\phi(\theta) = \lim_{n \to \infty} \varphi\left(\frac{\theta}{2^{n}}\right)^{2^{n}}\prod_{k=0}^{n-1}\Phi\left(\frac{\theta}{2^{k}}\right)^{2^{k}}
% \]
is a fixed point whenever this limit exists.

@@

The crucial step in the derivation of the infinitesimal model in \cite{barton2017infinitesimal} is to show that, following the arguments in \cite{Fisher1918}, knowing an individual's trait value gives vanishingly little information about the contribution of any given locus as $M \to \infty$, which in turn allows one to show that the within-family Mendelian sampling term is independent of the parental traits. Here, we will consider how conditioning an individual's trait value affects the law of a single allelic contribution in the Cauchy setting described above. A similar argument can be made for any $\alpha$-stable distribution, but the lack of explicit probability density functions for generic $\alpha \in (0,2)$ adds complexity to the argument without adding any insight.

@@

\begin{figure}
    \begin{center}
        \includegraphics{sims/neutral_seg_noise}
    \end{center}
    \caption{
        The distribution of Mendelian sampling terms
        with Gaussian \textbf{(A-C)} and Cauchy \textbf{(D-F)} effect size distributions,
        from the same simulations shown in Figure~\ref{fig:trait_distrns},
        for all individuals born during the last 1,000 time units of the simulation.
        \textbf{(A,D)} Mendelian noise (offspring trait minus midparent) plotted against midparent.
        \textbf{(B,E)} Histograms of the Mendelian sampling term.
        \textbf{(C,F)} Enrichment in quantiles of the Mendelian sampling term conditional on midparent value.
        To compute the latter,
        let $q_0, q_2, q_4, \ldots, q_{98}, q_{100}$ divide the distribution of the Mendelian sampling term
        into 50 bins with roughly equal numbers in each (i.e., the quantiles),
        let $p_k(a,b)$ denote the proportion of the offspring of parents with midparent value
        is in $[a, b)$ whose Mendelian sampling term is in $[q_{k-1}, q_k)$,
        and let $p_k$ be the proportion of all offspring with Mendelian noise in $[q_{k-1}, q_k)$
        (so, $p_k \approx 0.02$).
        Then, each line shows $p_k(a,b)/p_k$
        plotted against the midpoint of the corresponding quantile bin $(q_{k-1},q_k)$,
        for $(a,b)$ chosen to span the three segments of 5\% of
        ``low'', ``medium'', and ``high'' midparents, i.e.:
        midparents between (red) the 7.5\% and 12.5\% quantiles,
        (black) the 47.5\% and 52.5\% quantiles, and
        (green) the 87.5\% and 92.5\% quantiles of midparent value.
        \label{fig:seg_noise}
    }
\end{figure}

@@

The infinitesimal model is a powerful and tractable model of trait inheritance,
but is polygenic to a fault,
excluding the possibility of loci of large effect
(that are often seen in practice).
In this paper,
we have looked at the possibility of using a broader range of central limit theorems
to extend the infinitesimal model to traits for which 
effect size distributions fall in the domain of attraction of the $\alpha$-stable distributions
(i.e., a larger proportion of variance
is explained by a smaller number of loci).
Indeed, we found evidence that this is true of many real traits,
although our relatively uncritical use of GWAS estimates makes this evidence suggestive at best.
Next, we show that in fact the infinitesimal model \emph{cannot} describe non-Gaussian distributions
\comment{XXX TODO}.
A recurring theme in the rest of the paper is
that just as the infinitesimal model is useful precisely because it ignores the genetic details,
it cannot apply once some genetic details are important
(i.e., there are loci of large effect).
Also, we show that removing big effects and using the infinitesimal model for the remainder is good \comment{todo incorporate this}.
